Legion: Expressing Locality and Independence with Logical Regions

### 2012-04-27 doc/sc2012/circuit.tex

%\include{code_ex}

In this section we give an informal overview of Legion through an example.
Listing~\ref{lst:code_ex} shows code for an electrical
circuit simulation, which illustrates the core features of the programming model.

The circuit simulation takes a collection of
of wires and nodes where wires meet.  
The simulation repeatedly updates
currents, distributes charges, and updates voltages for as many
time steps as the simulation demands.  We first
describe the partitioning of regions in Section~\ref{subsec:partitioning}
and then describe the data type declarations 
and the constraints they enforce 
in Section~\ref{subsec:datatypes}.

@@

Line 18 declares the main simulator function, which takes a {\tt
Circuit} to be simulated.  This function specifies the regions it 
accesses and the {\em privileges} and {\em coherence} it requires of those regions.
In this case, the {\tt RWE} annotation specifies that the function
accesses the regions {\tt c.r\_all\_nodes} and {\tt c.r\_all\_wires}
with read-write privileges and {\em exclusive} coherence (i.e., no other
function can access the regions concurrently).  The simulation 
reads and writes all nodes and wires, and it must be done
with exclusive access to ensure correct results.  Privileges specify what
the function can do with the regions; coherence specifies what other
functions can do with the regions concurrently.  Functions that
declare their accessed regions, privileges, and coherence are called {\em tasks}
and are the unit of parallel execution in Legion.

Lines 19-27 are responsible for partitioning the circuit into {\tt MAX\_PIECES}
pieces that can be worked on in parallel.  To partition a region, we
provide a {\em coloring}, which is a relation between the elements
of a region and a set of colors.  A {\em partition} is an object which given
a coloring and a region $r$ contains for each color $c$ a {\em subregion} of $r$
containing all the elements of $r$ colored $c$.

Wires are simply partitioned in {\tt MAX\_PIECES} subregions (line 20).  
The nodes are first partitioned into 
{\em private} and {\em shared} nodes (line 21).  Private nodes
are touched by a single task in a phase, while shared nodes may
be referenced by multiple tasks in a phase.  The thick lines in Figure~\ref{sfig:part_fig:pvs}
show how the nodes in a small graph might be partitioned into three subsets.  Gray
nodes are shared and black nodes are private.  Note that
this partition is disjoint since each node has one color.
%Our current implementation uses small
%integers or booleans for the names of colors, but any enumerable set would be
%fine.  The coloring relation need not be total - it can leave out some elements entirely.  And although it is not required to be injective (it can map a single
%element to multiple colors), the fact that injective mappings will always
%result in disjoint subregions makes them strongly encouraged (when possible).
%Our example uses four colorings to create circuit partitions.

We next further partition the private and shared subregions into {\tt
MAX\_PIECES} disjoint subregions each (lines 23-24); one subregion of the private and shared node partitions
is shown in Figures~\ref{sfig:part_fig:p_i} and \ref{sfig:part_fig:s_i} respectively.  
We partition the shared nodes again
into the sets of ghost nodes for each task.  Since ghost nodes may be read by multiple tasks,
nodes may be colored multiple times and the ghost regions are not disjoint; one of the ghost subregions is shown in Figure~\ref{sfig:part_fig:g_i}.
Figure~\ref{sfig:part_fig:tree} shows the final hierarchy of node partitions.  The $*$ symbol indicates a partition is 
disjoint.

@@

Lines 36-40 form the bulk of the actual simulation, performing three passes
over the circuit in each iteration.  Each pass loops over an array of pieces (constructed on lines 30-34 from the partitions),
spawning a task for each piece.   There are no explicit requests for
parallel execution, nor is there synchronization required between the
passes.  Both the fact that the tasks can be run in parallel within a pass
and the required inter-pass dependencies are determined automatically
by the Legion runtime based on the region access annotations on the task declarations
(see Section~\ref{sec:soop}).

@@

\subsection{Data Types and Constraints}
\label{subsec:datatypes}
The two basic data types are {\tt Node}s and {\tt Wire}s.
{\tt Node}s are linked in circular lists.
The simulation relies on different lists being in different
regions, so the {\tt Node} definition on line 1 is parameterized by a region
{\tt rn}.  The {\tt next} field of a $\tt Node \langle rn \rangle$ has
type $\tt Node \langle rn \rangle \mbox{\tt @} rn$, which means the field points
to an object in region {\tt rn} of type $\tt Node \langle rn \rangle$, 
implying that all nodes of the list are in region {\tt rn}.

{\tt Wire}s are similar, except that a wire refers to two nodes to
which it is electrically connected.  The two {\tt Node}s are in region {\tt rn},
but they may point to {\tt Node}s in a different region {\tt rn2}.  
This is a common pattern in Legion code, and
arises when a constraint needs to be placed on the objects directly
referenced, and a second (presumably weaker) constraint on objects
that are indirectly reached through the direct pointers.

In cases where we need to create compound data structures with types whose
region parameters are self-referential we declare a {\em region relationship}.
A region relationship is a structure with named fields, 
except that any region fields can be used in the type declarations
for other fields.  Lines 4-9 declare the region relationship {\tt Circuit}.
In {\tt Circuit} the region {\tt r\_all\_nodes} contains {\tt
Node}s that point only into region {\tt r\_all\_nodes}, guaranteeing that
all {\tt Node}s are in region {\tt r\_all\_nodes}.  (Note that regions are
declared with the single data type they can contain.)  Similarly, the
type of the wire region's element not only guarantees that every wire
in the circuit is in the wire region, but also guarantees that every
node referred to by any wire is in the circuit's node region.

Since a region relationship can have
fields with types that refer to its own fields' regions, it is often not possible
to assign fields one at a time without violating the type checking rules.
The simultaneous assignment operator asks the type checker only to ensure
that fields have the correct types after all the fields are updated.  In
many cases, the target type also tries to describe tighter constraints on
the region parameters used for field types.  The {\tt region\_cast} operator
allows the programmer to assert the correctness of this transformation, which
can be verified with a linear scan over the elements of a region.
Lines 32-34 illustrate a simultaneous
assignment to update a {\tt CircuitPiece} region relationship.

The last type definition is for the pieces into which the circuit is partitioned.
The original node and wire regions are specified as parameters, and the
four kinds of subregions that are created are made fields of the
{\tt CircuitPiece} (lines 10-17).  This declaration also illustrates {\em subregion constraints}
placed on the three region fields stating the region fields must be subregions of the 
node and wire regions provided as parameters (line 16).  Since the simulation 
iterates over the nodes owned by a {\tt CircuitPiece}, the parameterization of the {\tt rn\_pvt} (private nodes),
{\tt rn\_shr} (shared nodes), and {\tt first\_node} fields constrains the linked list of
nodes to reside entirely in those two regions.  The linked list of wires is
similarly constrained, and the wire's node pointers are constrained to fall
into one of the three subregions.
% defined in the {\tt CircuitPiece}.  
Finally,
the use of {\tt rn} for {\tt rn\_ghost}'s parameter and the second parameter
in the wire type places no constraints on which nodes a ghost node might point
to.

>>>>

In this section we give an overview of Legion using an example.
Listing~\ref{lst:code_ex} shows code for an electrical
circuit simulation, which illustrates the core features of the programming model.

The circuit simulation takes a collection of
of wires and nodes where wires meet.  
The simulation repeatedly updates
currents, distributes charges, and updates voltages for as many
time steps as the simulation demands.  We first
describe the partitioning of regions in Section~\ref{subsec:partitioning}
and then describe the data type declarations in Section~\ref{subsec:datatypes}.

@@

Line 9 declares the main simulator function, which takes a {\tt
Circuit} to be simulated.  This function specifies the regions it 
accesses and the {\em privileges} and {\em coherence} it requires of those regions.
In this case, the {\tt RWE} annotation specifies that the function
accesses the regions {\tt c.r\_all\_nodes} and {\tt c.r\_all\_wires}
with read-write privileges and {\em exclusive} coherence (i.e., no other
function can access these two regions concurrently).  The simulation 
reads and writes all nodes and wires, and it must be done
with exclusive access to ensure correct results.  Privileges specify what
the function can do with the regions; coherence specifies what other
functions can do with the regions concurrently.  Functions that
declare their accessed regions, privileges, and coherence are called {\em tasks}
and are the unit of parallel execution in Legion.

Lines 15-21 are responsible for partitioning the circuit into {\tt MAX\_PIECES}
pieces that can be worked on in parallel.  To partition a region, we
provide a {\em coloring}, which is a relation between the elements
of a region and a set of colors.  A {\em partition} is an object which given
a coloring and a region $r$ contains for each color $c$ a {\em subregion} of $r$
containing all the elements of $r$ colored $c$.

Wires are simply partitioned in {\tt MAX\_PIECES} subregions (line 16).  
The nodes are first partitioned into 
{\em private} and {\em shared} nodes (line 21).  Private nodes
are touched by a single task in a phase, while shared nodes may
be referenced by multiple tasks in a phase.  The thick lines in Figure~\ref{sfig:part_fig:pvs}
show how the nodes in a small graph might be partitioned into three subsets.  Gray
nodes are shared and black nodes are private.  Note that
this partition is disjoint since each node has one color.
%Our current implementation uses small
%integers or booleans for the names of colors, but any enumerable set would be
%fine.  The coloring relation need not be total - it can leave out some elements entirely.  And although it is not required to be injective (it can map a single
%element to multiple colors), the fact that injective mappings will always
%result in disjoint subregions makes them strongly encouraged (when possible).
%Our example uses four colorings to create circuit partitions.

We next further partition the private and shared subregions into {\tt
  MAX\_PIECES} disjoint subregions each (lines 20-23); one subregion
of the private and shared node partitions is shown in
Figures~\ref{sfig:part_fig:p_i} and \ref{sfig:part_fig:s_i}
respectively.

We partition the shared nodes a second time using a
different coloring into the sets of ghost nodes for each task.  This 
partition is noteworthy because the subregions are {\em not} disjoint:
ghost nodes may be read by multiple tasks, and thus a node $n$  may belong 
to more than one ghost region, which corresponds to $n$ being assigned
more than one color by the coloring {\tt node\_neighbor\_map}.
One of the ghost subregions is shown in Figure~\ref{sfig:part_fig:g_i}.

Figure~\ref{sfig:part_fig:tree} shows the final hierarchy of node
partitions;  the $*$ symbol indicates a partition is disjoint. This
{\em region tree} data structure plays an important role in scheduling
tasks for out-of-order execution (see Section~\ref{sec:soop}).

@@

Lines 31-57 form the bulk of the actual simulation, performing three passes
over the circuit in each iteration.  Each pass loops over an array of pieces (constructed on lines 27-30 from the partitions),
spawning a task for each piece.   There are no explicit requests for
parallel execution (the default semantics of {\tt spawn} is sequential 
execution of the spawned tasks) nor is there synchronization required between the
passes.  Both the fact that the tasks can be run in parallel within a pass
and the required inter-pass dependencies are determined automatically
by the Legion runtime based on the region access annotations on the task declarations
(see Section~\ref{sec:soop}).

@@

\subsection{Data Types}
\label{subsec:datatypes}
The two basic data types in the circuit simulation are {\tt Node}s and
{\tt Wire}s declared on lines 1 and 2 of Listing~\ref{lst:code_ex}.
As is standard in region-based systems, allocating a {\tt Node}, {\tt
  Wire}, or any value in the heap requires naming the logical region
in which the value is allocated.  Furthermore, pointers to heap values
declare the region to which they point.  For example, the definition
of {\tt Wire} is parameterized on the region {\tt rn} to which the
{\tt Node} pointers in fields {\tt in\_nodes} and {\tt out\_nodes}
point, which guarantees that the two endpoints of a wire are always in
the same region.

Regions are first-class values in Legion and can be stored in heap
data structures.  For example, a {\tt Circuit} (defined on line 3)
holds the regions hold all nodes and wires of the circuit and a {\tt
  CircuitPiece} (defined on line 4) hold pointers to the private,
shared, and ghost nodes as well as the private wires of its piece of
the circuit.
